---
title: "Blood donation solution using Neural Networks"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description
Use information about each donor's history:

- Months since Last Donation: this is the number of monthis since this donor's most recent donation.
- Number of Donations: this is the total number of donations that the donor has made.
- Total Volume Donated: this is the total amound of blood that the donor has donated in cubuc centimeters.
- Months since First Donation: this is the number of months since the donor's first donation.

## Problems

1. BUILD A PREDICTING MODEL BASED ON DONOR'S HISTORY AND OUTCOME OF MADE DONATION IN MARCH 2007
2. USING MODEL BUILT IN PROBLEM 1, PREDICT THE OUTCOME OF DONATIONS IN NEXT BLOOD DONATION CAMPAIGN (Assume that next donation will be on May 2007)

## Solution
```{r results = 'hide', echo = FALSE, message = FALSE, warning = FALSE}
# Load necessary packages
library(neuralnet)
library(tidyr)
library(dplyr)
library(gmodels)
library(ggplot2)
library(cowplot)
```
### Load and prepare blood donation dataset

Load the training dataset into data frame `bd` 
```{r}
bd <- read.csv('blood_donation.csv')
head(bd)
str(bd)
```

### Explore the distributed single variable plot

```{r echo=FALSE}
jitter <- ggplot(bd, aes(factor(Made.Donation.in.March.2007), Number.of.Donations, col = factor(Made.Donation.in.March.2007))) + geom_jitter() + theme(legend.position="none")
boxplot <- ggplot(bd, aes(factor(Made.Donation.in.March.2007), Number.of.Donations, col = factor(Made.Donation.in.March.2007))) + geom_boxplot() + theme(legend.position="none")
plot_grid(jitter, boxplot, ncol = 2, nrow = 1)
```
```{r echo=FALSE}
jitter <- ggplot(bd, aes(factor(Made.Donation.in.March.2007), Total.Volume.Donated..c.c.., col = factor(Made.Donation.in.March.2007))) + geom_jitter() + theme(legend.position="none")
boxplot <- ggplot(bd, aes(factor(Made.Donation.in.March.2007), Total.Volume.Donated..c.c.., col = factor(Made.Donation.in.March.2007))) + geom_boxplot() + theme(legend.position="none")
plot_grid(jitter, boxplot, ncol = 2, nrow = 1)
```
```{r echo=FALSE}
jitter <- ggplot(bd, aes(factor(Made.Donation.in.March.2007), Months.since.First.Donation, col = factor(Made.Donation.in.March.2007))) + geom_jitter() + theme(legend.position="none")
boxplot <- ggplot(bd, aes(factor(Made.Donation.in.March.2007), Months.since.First.Donation, col = factor(Made.Donation.in.March.2007))) + geom_boxplot() + theme(legend.position="none")
plot_grid(jitter, boxplot, ncol = 2, nrow = 1)
```
```{r echo=FALSE}
jitter <- ggplot(bd, aes(factor(Made.Donation.in.March.2007), Months.since.Last.Donation, col = factor(Made.Donation.in.March.2007))) + geom_jitter() + theme(legend.position="none")
boxplot <- ggplot(bd, aes(factor(Made.Donation.in.March.2007), Months.since.Last.Donation, col = factor(Made.Donation.in.March.2007))) + geom_boxplot() + theme(legend.position="none")
plot_grid(jitter, boxplot, ncol = 2, nrow = 1)
```

### Explore the 2 variables combination plot

```{r echo=FALSE}
j1 <- ggplot(bd, aes(Number.of.Donations, Total.Volume.Donated..c.c.., col = factor(Made.Donation.in.March.2007))) + geom_jitter() + theme(legend.position="none")
j2 <- ggplot(bd, aes(Number.of.Donations, Months.since.First.Donation, col = factor(Made.Donation.in.March.2007))) + geom_jitter() + theme(legend.position="none")
j3 <- ggplot(bd, aes(Number.of.Donations, Months.since.Last.Donation, col = factor(Made.Donation.in.March.2007))) + geom_jitter() + theme(legend.position="none")
j4 <- ggplot(bd, aes(Months.since.First.Donation, Months.since.Last.Donation, col = factor(Made.Donation.in.March.2007))) + geom_jitter() + theme(legend.position="none")
plot_grid(j1, j2, j3, j4, ncol = 2, nrow = 2)
```

**Summary:**

- Two variables `Number.of.Donations` and `Total.Volume.Donated..c.c..` strictly linearly dependent to each other so we could remove one of these out of dataset without decreasing effectiveness of our predicting model
- Since the data points are not linearly separated, we should introduce some features to help the predicting model, namely `DonationPerMonth` and `MonthPerDonation` computed as follow:

$$DonateDuration = Months.since.First.Donation - Months.since.Last.Donation$$
$$DonationsPerMonth = \frac{Number.of.Donations}{DonateDuration}$$
$$MonthsPerDonation = \frac{DonateDuration}{Number.of.Donations}$$

### Prepare the training dataset

```{r}
prepare <- function (data) {
  data %>%
  rename(MonthsLD = Months.since.Last.Donation,
         NumberDonations = Number.of.Donations,
         TotalVolume = Total.Volume.Donated..c.c..,
         MonthsFD = Months.since.First.Donation,
         March.2007 = Made.Donation.in.March.2007) %>%
  # mutate(March.2007 = as.factor(March.2007)) %>%
  mutate(DonateDuration = MonthsFD - MonthsLD + 1) %>%
  mutate(DonationsPerMonth = NumberDonations / DonateDuration) %>%
  mutate(MonthsPerDonation = DonateDuration / NumberDonations) %>%
  select(NumberDonations,
         MonthsFD,
         MonthsLD,
         # TotalVolume,
         # DonateDuration,
         DonationsPerMonth,
         MonthsPerDonation,
         March.2007)
}
bd.prepared <- prepare(bd)
head(bd.prepared)
str(bd.prepared)
```

Split the original training set into training and validation sets
```{r}
n <- nrow(bd.prepared)
s <- sample(1:n, 2*n/3)
bd.train <- bd.prepared[s,]
head(bd.train)
bd.validate <- bd.prepared[-s,]
head(bd.validate)
```

#### Logistic model

First we will train a simple logistic model first to compare with neural network later on
```{r}
bd.logis <- glm(data = bd.train,
                March.2007 ~ NumberDonations + MonthsFD + MonthsLD + DonationsPerMonth + MonthsPerDonation,
                family = binomial)

summary(bd.logis)

bd.logis.predict <- predict(bd.logis, bd.validate, type = "response")

CrossTable(x = bd.validate$March.2007, y = round(bd.logis.predict), prop.chisq = TRUE)
```

#### Neural networks

```{r}
bd.nn <- neuralnet(March.2007 ~ NumberDonations + MonthsFD + MonthsLD + DonationsPerMonth + MonthsPerDonation,
                   bd.train, act.fct = "logistic", hidden = c(3), threshold = 0.01, stepmax = 10000)

summary(bd.nn)

bd.nn.predict <- neuralnet::compute(bd.nn, select(bd.validate,
                                                  NumberDonations,
                                                  MonthsFD,
                                                  MonthsLD,
                                                  DonationsPerMonth,
                                                  MonthsPerDonation))

bd.nn.predict.result <- round(bd.nn.predict$net.result)

CrossTable(x = bd.validate$March.2007, y = bd.nn.predict.result, prop.chisq = TRUE)
```

## Testing 

Load and prepare the testing dataset into data frame `bd.test`
```{r}
bd.test <- read.csv('blood_donation_test.csv')
bd.test <- prepare(bd.test)
head(bd)
str(bd)
```

Test the logistic model
```{r}
bd.logis.test <- predict(bd.logis, bd.test, type = "response")

CrossTable(x = bd.test$March.2007, y = round(bd.logis.test), prop.chisq = TRUE)
```

Compare with neural networks
```{r}
bd.nn.test <- neuralnet::compute(bd.nn, select(bd.test,
                                               NumberDonations,
                                               MonthsFD,
                                               MonthsLD,
                                               DonationsPerMonth,
                                               MonthsPerDonation))

bd.nn.test.result <- round(bd.nn.test$net.result)

CrossTable(x = bd.test$March.2007, y = bd.nn.test.result, prop.chisq = TRUE)
```

## Problem 2: using model built in problem 1, predict the outcome of donations in next blood donation campaign

Modify the original `bd` dataset to reflect accumulative data at May 2007

- If person donate at March 2007 (`Made.Donation.in.March.2007` equal 1) then his/her `Number.of.Donations` will increase by 1 and `Months.since.Last.Donation` will be reset to 2
- If person not join the donation at March 2007 then his/her `Number.of.Donations` remain the same and `Months.since.Last.Donation` will increase by 2

```{r}
bd2 <- bd %>%
  mutate(Months.since.First.Donation = Months.since.First.Donation + 2) %>%
  mutate(Months.since.Last.Donation = if_else(Made.Donation.in.March.2007 == 1,
                                              2, Months.since.Last.Donation + 2)) %>%
  mutate(Number.of.Donations = Number.of.Donations + Made.Donation.in.March.2007) %>%
  prepare()
```

Predict the outcome of donation at May 2007 using the previous neural networks
```{r}
bd2.predict <- neuralnet::compute(bd.nn, select(bd2,
                                                NumberDonations,
                                                MonthsFD,
                                                MonthsLD,
                                                DonationsPerMonth,
                                                MonthsPerDonation))

bd2.predict.df <- data.frame(probability = bd2.predict$net.result,
                             prediction = round(bd2.predict$net.result))
summary(bd2.predict.df)
```



